---
title: "Stat 154 HW 01"
author: "Mokssh Surve"
date: "2/3/2020"
output: rmarkdown::github_document
---

```{r}

library('ramify')

```

# Problem 5
## 5a) - inner_prod()
```{r}

inner_prod <- function(a,b) {
  
  if(length(a) != length(b))
  { stop("Vector lengths must be the same") }
  
  prod <- as.numeric(t(a) %*% b)
  prod
}

y <- c(1, 1, 1, 1, 1)
z <- c(2, 4, 6, 8, 10)

inner_prod(y,y)
inner_prod(z,z)
inner_prod(y,z)

```

## 5b) - vnorm()
```{r}

vnorm <- function(a) {
  
  eu_norm <- sqrt(inner_prod(a,a))
  eu_norm
}

b <- c(1,2,3,4,5)
vnorm(b)

```

## 5c) - unit_norm()
```{r}

unit_norm <- function(a) {
  
  unit <- a / vnorm(a)
  unit
}

x <- c(1,2,3)
y <- c(1,1,1)
z <- c(1,0,0)

unit_norm(x)
unit_norm(y)
unit_norm(z)

```

## 5d) - vector_proj()
```{r}

vector_proj <- function(a,b) {
  
  if(length(a) != length(b))
  { stop("Vector lengths must be the same") }
  
  vec_proj <- (inner_prod(a,b)/inner_prod(b,b)) * b
  vec_proj
}

x <- c(1,2,3)
y <- c(1,1,1)

vector_proj(x,y)

```

## 5e) - scalar_proj()
```{r}

scalar_proj <- function(a,b) {
  
  if(length(a) != length(b))
  { stop("Vector lengths must be the same") }
  
  scal_proj <- inner_prod(a,b) / vnorm(b)
  scal_proj
}

x <- c(1,2,3)
y <- c(1,1,1)

scalar_proj(x,y)

```


# Problem 6
## 6a) Implementing Power Method 
```{r}

power_method <- function(A, n) {
  
  # initialising the initial random vector
  rows <- nrow(A)
  w_0 <- randn(rows,1)

  w_old <- w_0
  
  for(k in 1:n)
  {
    w_new <- (A %*% w_old) 
    s_new <- max(abs(w_new))
    w_new <- w_new / s_new
    
    # appending new column to matrix of approximated eigenvectors
    w_0 <- cbind(w_0, w_new)
   
    w_old <- w_new
  }

  # converting vector so as to have unit norm
  w_new <- unit_norm(w_new)
  
  #creating list to output dominant eigen vector w_k+1 and eigen value
  out <- list(dom_vec=w_new, dom_value=s_new)
  out
  
}

A <- matrix(c(5,-4,3,-14,4,6,11,-4,-3), nrow=3, ncol=3)
power_method(A, 10)
eigen(A)

#eigenvalue given by Rayleigh Quotient (4)
vec <- power_method(A, 10)$dom_vec
rayleigh <- (t(vec) %*% A %*% vec) / (t(vec) %*% vec)
rayleigh

#taking unit_norm() (5)
unit_norm(vec)
#checking if euclidean norm is 1
vnorm(vec)

```

From this, it can be seen that the dominant (largest) eigenvalues & eigenvectors are almost the same for both functions. \newline
• **power_method()** outputs 12.05289 \newline
• **eigen() outputs** 12 \newline
• **Raleigh Quotient** gives us 11.97889 \newline
*Note: The values (except for eigen()) might be a bit different since each time the function is run, a pseudorandom vector is created*

## 6b) Other Scaling Options
```{r}

#A - Matrix
#n - no. of iterations
#p - L_p norm parameter
power_method_p <- function(A, n, p) {
  
  # initialising the initial random vector
  rows <- nrow(A)
  w_0 <- randn(rows,1)
  w_old <- w_0
  
  for(k in 1:n)
  {
    w_new <- (A %*% w_old) 
    s_new <- (sum(abs(w_new))) ^(1/p)
    w_new <- w_new / s_new
    
    # appending new column to matrix of approximated eigenvectors
    w_0 <- cbind(w_0, w_new)
   
    w_old <- w_new
  }
  
  # converting vector so as to have unit norm
  w_new <- unit_norm(w_new)
  
  out <- list(dom_vec=w_new, dom_value=s_new)
  out
  
}
  
power_method_p(A, 20, 2)
eigen(A)


```

From this, it can be confirmed that the dominant (largest) eigenvalues & eigenvectors are almost the same for both functions. \newline
• **power_method()** outputs 11.99988 \newline
• **eigen()** outputs 12 \newline
*Note: The values (except for eigen()) might be a bit different since each time the function is run, a pseudorandom vector is created*


## 6c) Deflation & More eigenvectors
```{r}

B <- cbind(c(5, 1), c(1, 5))

#applying Power Method to get first eigenvector & eigenvalue
v_1 <- power_method(B, 20)$dom_vec
lambda_1 <- power_method(B, 20)$dom_value


#deflating matrix B
B_1 <- B - lambda_1 * ((v_1) %*% t(v_1))

#getting the next eigenvector & eigenvalue
v_2 <- power_method(B_1, 20)$dom_vec
lambda_2 <- power_method(B_1, 20)$dom_value

#displaying first 2 eigenvectors
cbind(v_1, v_2)

#displaying their respective eigenvalues
cbind(lambda_1, lambda_2)

eigen(B)
```

Thus, it can be confirmed, that Deflation works as eigen() gives the same eigenvalues & eigenvectors.








